Probabilistic Forecasting of the Disturbance Storm Time Index: An Autoregressive Gaussian Process approach

### 2017-05-26 revised-draft.tex
Added section on Akasofu epsilon

    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to the weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}. 
\end{enumerate}

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. $Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}. Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$.

In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.

>>>>

    \item $Dst$: A continuous hourly index which gives a measure of the weakening or strengthening of the Earth's equatorial magnetic field due to particle injection in the magnetosphere. Particle injection has a number of sources such as, weakening or strengthening of the ring currents and the geomagnetic storms \citep{DesslerAndParker}, near Earth cross tail current \citep{ganushkina2004long} and \citep{angeo-28-123-2010}, partial ring current \citep{JGRA:JGRA15878}, substorm current wedge \citep{JGRA:JGRA15211}, magnetopause current, etc . 
\end{enumerate}

@@

%Talk about NARMAX Dst
Another important empirical model used to predict $Dst$ is the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations of monomials to include in the refined model by using a criterion called the \emph{error reduction ratio} (ERR). The parameters of the so called NARMAX OLS-ERR model are calculated by solving the \emph{ordinary least squares} (OLS) problem arising from a quadratic objective function. It must be noted that the NARMAX methodology is not limited to polynomial functions, rather any set of basis function expansions can be used with it, such as but not limited to radial basis functions, wavelets etc \citep{doi:10.1080/00207720600903011}, \citep{JGRA:JGRA17327}. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

Auto-regressive models can be augmented by including exogenous quantities in the inputs $\mathbf{x}_{t-1}$ at each time step, in order to improve predictive accuracy. While modeling $Dst$ using the OMNI data, one must choose which solar wind quantities to include in the exogenous inputs of the predictive model. This choice is not straight forward and eventually requires a compromise between including important solar wind quantities and keeping the input space managable in the interest of simplicity.

\subsubsection{Choice of Solar Wind Inputs}

$Dst$ gives a measure of ring currents, which are modulated by plasma sheet particle injections into the inner magnetosphere during sub-storms. Studies have shown that the substorm occurrence rate increases with solar wind velocity (high speed streams) \citet{Kissinger2011,Newell2016}. Prolonged southward interplanetary magnetic field (IMF) $z$-component ($B_z$) is needed for sub-storms to occur \citet{McPherron1986}. An increase in the solar wind electric field, $V_{sw}B_z$, can increase the dawn-dusk electric field in the magnetotail, which in turn determines the amount of plasma sheet particle that move to the inner magnetosphere \citet{Friedel2001}.

Apart from $V$ and $B_z$, other quantities which have been shown to correlate with geomagnetic activity are solar wind dynamic pressure $P$, clock angle $tan \theta = \frac{B_y}{B_z}$, Akasofu $\epsilon$ \citep{1986AkasofuE} and solar wind magnetosphere coupling functions \citep{JGRA:JGRA21451}.

Although solar wind magnetospheric coupling functions have been shown to have correlation with geomagnetic indices, they are expressible in terms of $V$ and $B_z$, and hence we do not need to include them as explicit inputs to the model. \emph{Gaussian Process} models derive their strength from automatic feature construction achieved by the covariance functions (interested readers may refer to chapters 6 and 7  in \cite{Rasmussen:2005:GPM:1162254}). Therefore as long as coupling functions can be approximated in the \emph{eigenspace} of the covariance function we need not make them explicit in the input features.

Therefore, our exogenous parameters consist of solar wind velocity $V_{sw}$ and IMF $B_z$. In this model we choose distinct time lags $p$, $p_{v}$ and $p_{b}$ for $Dst$, $V$ and $B_z$ respectively.
